treats manual

### 2023-11-21 inst/gitbook/_book/treats_manual.tex
fully updated manual

\title{treats manual}
\author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}})}
\date{2023-05-25}

@@

@@

@@

\begin{verbatim}
##  ---- treats object ---- 
## Birth death process with modifiers:
## speciation: 1.
## extinction: 0.2.
## No modifiers applied to the branch length, selection and speciation processes (default).
## 
## Simulated phylogenetic tree (x$tree):
## 
## Phylogenetic tree with 24 tips and 23 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
## 
## Simulated trait data (x$data):
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

Note however, that the printing information is now updated to state that you've added a \texttt{"modifiers"} object (even though it's the same as the default).

For more interesting simulations however, you can provide modifiers that actually modify the birth-death process.
We can create one for example that makes species go extinct if their ancestor has a negative trait value.
For that we need to create a \texttt{"modifiers"} object that modifies the \texttt{speciation} process with a specific condition and a specific modification when that condition is met.
For a speciation to occur (and a species to not go extinct), the algorithm draws a random value \emph{x} between 0 and 1 and if the value is smaller than the speciation parameter divided by the speciation + extinction parameter, the lineage speciates (\(x < (\lambda \ (\lambda + \mu))\)), else it goes extinct.

Here we can create a modifier forcing the lineage to go extinct every time a specific condition is met, e.g.~if the lineage trait value is negative.
In other words, if the lineage's trait value is negative, make sure that \emph{x} is equal to 1, so that the speciation condition \(x < (\lambda \ (\lambda + \mu))\) is never met.

@@

We can then provide these two functions (the condition \texttt{negative.ancestor} and how to modify the speciation event when this condition is met \texttt{going.extinct}).
If you are an advances \texttt{treats} user, you can design your own \texttt{speciation} function but if you just want to use a normal \texttt{speciation} function, you can use the default one from \texttt{treats} called \texttt{speciation}.

@@

\begin{verbatim}
##  ---- treats object ---- 
## Birth death process with modifiers:
## speciation: 1.
## extinction: 0.2.
## Default branch length process.
## Default selection process.
## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (going.extinct).
## 
## Simulated phylogenetic tree (x$tree):
## 
## Phylogenetic tree with 22 tips and 21 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
## 
## Simulated trait data (x$data):
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

@@

\begin{verbatim}
##  ---- treats birth-death parameters object ---- 
## speciation: 0.352330414024189.
## extinction: 0.287616664509542.
\end{verbatim}

@@

Et voila! We now have a distribution of 50 trees and 50 datasets from which we can calculate disparity.
We can convert this list into \texttt{dispRity} using the \texttt{treats.dispRity} function.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Converting the data to be passed to dispRity functions}
\NormalTok{dispRity\_data \textless{}{-}}\StringTok{ }\KeywordTok{dispRitreats}\NormalTok{(sim\_data)}

\CommentTok{\#\# Creating the time series (by default dispRitreats scales the trees to be 1 unit of times old)}
\NormalTok{time\_series \textless{}{-}}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ dispRity\_data}\OperatorTok{$}\NormalTok{data,}
                              \DataTypeTok{tree =}\NormalTok{ dispRity\_data}\OperatorTok{$}\NormalTok{tree,}
                              \DataTypeTok{time =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \FloatTok{0.9}\NormalTok{, }\DataTypeTok{to =} \DecValTok{0}\NormalTok{, }\DataTypeTok{by =} \FloatTok{{-}0.1}\NormalTok{),}
                              \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"proximity"}\NormalTok{)}

\CommentTok{\#\# Calculating the disparity from the simulated data}
\NormalTok{simulated\_disparity \textless{}{-}}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(time\_series, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
                                \DataTypeTok{verbose =} \OtherTok{TRUE}\NormalTok{)}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(simulated\_disparity, }\DataTypeTok{main =} \StringTok{"simulated disparity"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(change\_in\_size, }\DataTypeTok{main =} \StringTok{"observed disparity"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Creating a random mass extinction}
\NormalTok{random\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}\DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
                                 \DataTypeTok{condition =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{25}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{),}
                                 \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.75}\NormalTok{))}
\CommentTok{\#\# Creating an extrinction that removes species with negative trait values}
\NormalTok{negative\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}\DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
                                   \DataTypeTok{condition =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{25}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{),}
                                   \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DataTypeTok{x =} \DecValTok{0}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Converting the data to be passed to dispRity functions}
\NormalTok{rand\_ext\_data \textless{}{-}}\StringTok{ }\KeywordTok{dispRitreats}\NormalTok{(sim\_rand\_extinction)}
\NormalTok{trait\_ext\_data \textless{}{-}}\StringTok{ }\KeywordTok{dispRitreats}\NormalTok{(sim\_trait\_extinction)}

\CommentTok{\#\# Creating the time series}
\NormalTok{rand\_ext\_time\_series \textless{}{-}}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ rand\_ext\_data}\OperatorTok{$}\NormalTok{data,}
                                       \DataTypeTok{tree =}\NormalTok{ rand\_ext\_data}\OperatorTok{$}\NormalTok{tree,}
                                       \DataTypeTok{time =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \FloatTok{0.9}\NormalTok{, }\DataTypeTok{to =} \DecValTok{0}\NormalTok{, }\DataTypeTok{by =} \FloatTok{{-}0.1}\NormalTok{),}
                                       \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"proximity"}\NormalTok{)}
\NormalTok{trait\_ext\_time\_series \textless{}{-}}\StringTok{ }\KeywordTok{chrono.subsets}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ trait\_ext\_data}\OperatorTok{$}\NormalTok{data,}
                                        \DataTypeTok{tree =}\NormalTok{ trait\_ext\_data}\OperatorTok{$}\NormalTok{tree,}
                                        \DataTypeTok{time =} \KeywordTok{seq}\NormalTok{(}\DataTypeTok{from =} \FloatTok{0.9}\NormalTok{, }\DataTypeTok{to =} \DecValTok{0}\NormalTok{, }\DataTypeTok{by =} \FloatTok{{-}0.1}\NormalTok{),}
                                        \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{, }\DataTypeTok{model =} \StringTok{"proximity"}\NormalTok{)}

\CommentTok{\#\# Calculating the disparity from the simulated data}
\NormalTok{sim\_rand\_ext\_disp \textless{}{-}}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(rand\_ext\_time\_series, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
                              \DataTypeTok{verbose =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Calculating disparity..........Done.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sim\_trait\_ext\_disp \textless{}{-}}\StringTok{ }\KeywordTok{dispRity}\NormalTok{(trait\_ext\_time\_series, }\DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
                               \DataTypeTok{verbose =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Calculating disparity..........Done.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{2}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(change\_in\_size, }\DataTypeTok{main =} \StringTok{"observed disparity"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(simulated\_disparity, }\DataTypeTok{main =} \StringTok{"simulated disparity"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(sim\_rand\_ext\_disp, }\DataTypeTok{main =} \StringTok{"simulated disparity}\CharTok{\textbackslash{}n}\StringTok{(mass extinction)"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(simulated\_disparity, }\DataTypeTok{main =} \StringTok{"simulated disparity}\CharTok{\textbackslash{}n}\StringTok{(trait extinction)"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

Interestingly, with all our parameters used here and our specific disparity metric we cannot detcet the effect of a mass extinction.

@@

@@

\begin{verbatim}
## $main
## $main$A
## $main$A$process
## function (x0 = 0, edge.length = 1, Sigma = diag(length(x0)), 
##     ...) 
## {
##     return(t(MASS::mvrnorm(n = 1, mu = x0, Sigma = Sigma * edge.length, 
##         ...)))
## }
## <bytecode: 0x55b3b2541d88>
## <environment: namespace:treats>
## 
## $main$A$start
## [1] 0
## 
## $main$A$trait_id
## [1] 1
## 
## 
## 
## $background
## NULL
\end{verbatim}

@@

>>>>

\title{treats manual}
\author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}})}
\date{2023-11-21}

@@

\begin{verbatim}
##        --- dispRity package ---
## This is the CRAN release version (1.7.0) of the package.
## For news, vignettes and future releases,
## visit https://github.com/TGuillerme/dispRity
\end{verbatim}

@@

\hypertarget{treats-objects}{%
\subsection{\texorpdfstring{\texttt{"treats"} objects}{"treats" objects}}\label{treats-objects}}

\texttt{treats} will output either just a tree (class \texttt{"phylo"}) if no traits were generated or a \texttt{"treats"} object that contains both a \texttt{\$tree} (\texttt{"phylo"}) and a \texttt{\$data} (\texttt{"matrix"}) component.
Note that this \texttt{"treats"} class is generalised to most outputs of the package functions.
This allows for a smoother handeling of the objects outputs such as summarising the content of a \texttt{make.bd.params} output or visualising a trait output from \texttt{make.traits}.

@@

\begin{verbatim}
##  ---- treats object ---- 
## Birth death process with modifiers:
## speciation: 1.
## extinction: 0.2.
## No modifiers applied to the branch length, selection and speciation processes (default).
## Simulated phylogenetic tree (x$tree):
## 
## Phylogenetic tree with 24 tips and 23 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
## 
## Simulated trait data (x$data):
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

Note however, that the printing information is now updated to state that you've added a \texttt{"modifiers"} object (even though it's the same as the default).
In this specific case the \texttt{modifiers} object is not necessary as it is doing nothing different than when using the default one (i.e.~when no \texttt{modifiers} is provided).

For more interesting simulations however, you can provide modifiers that actually modify the birth-death process.
We can create one for example that makes species go extinct if their ancestor has a negative trait value.
For that we need to create a \texttt{"modifiers"} object that modifies the \texttt{speciation} process with a specific condition and a specific modification when that condition is met.
For a speciation to occur (and a species to not go extinct), the algorithm draws a random value \emph{x} between 0 and 1 and if the value is smaller than the speciation parameter divided by the speciation + extinction parameter, the lineage speciates (\(x < (\lambda \ (\lambda + \mu))\)), else it goes extinct.
\emph{x} is a random value that simulates some stochasticity in the evolutionary simulation.
At every step of the simulation, a different random value \emph{x} is drawn making sure that the resulting speciation or extinction is not predictable.

However, we can create a modifier that introduces less randomness (or a different level of randomness) by forcing the lineage to go extinct every time a specific condition is met, e.g.~if the lineage trait value is negative.
This can be done for example by not randomly generating the value \emph{x} but giving it a fixed value of say 1 so that the estimation of the speciation function \(x < (\lambda \ (\lambda + \mu))\) is always met.
In other words, if the lineage's trait value is negative, make sure that \emph{x} is equal to 1, so that the speciation condition \(x < (\lambda \ (\lambda + \mu))\) is never met.

@@

This \texttt{going.extinct} function is to be contrasted with the normal modifier for selecting the value \texttt{x} which is \texttt{runif(1)} (drawing a random value between 0 and 1).
We can then provide these two functions (the condition \texttt{negative.ancestor} and how to modify the speciation event when this condition is met \texttt{going.extinct}).
If you are an advances \texttt{treats} user, you can design your own \texttt{speciation} function but if you just want to use a normal \texttt{speciation} function, you can use the default one from \texttt{treats} called \texttt{speciation}.

@@

\begin{verbatim}
##  ---- treats object ---- 
## Birth death process with modifiers:
## speciation: 1.
## extinction: 0.2.
## Default branch length process.
## Default selection process.
## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (going.extinct).
## Simulated phylogenetic tree (x$tree):
## 
## Phylogenetic tree with 22 tips and 21 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
## 
## Simulated trait data (x$data):
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

@@

\begin{verbatim}
##  ---- treats birth-death parameters object ---- 
## speciation: 0.352.
## extinction: 0.288.
\end{verbatim}

@@

Et voila! We now have a distribution of 50 trees and 50 datasets from which we can calculate disparity.
We can directly pass the results to the \texttt{dispRity} package pipeline using the \texttt{dispRitreats} function:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Calculate the dispRity for all the simulations}
\NormalTok{simulated\_disparity \textless{}{-}}\StringTok{ }\KeywordTok{dispRitreats}\NormalTok{(sim\_data,}
                                    \DataTypeTok{method =} \StringTok{"continuous"}\NormalTok{,}
                                    \DataTypeTok{model  =} \StringTok{"proximity"}\NormalTok{,}
                                    \DataTypeTok{time   =} \DecValTok{10}\NormalTok{,}
                                    \DataTypeTok{metric =} \KeywordTok{c}\NormalTok{(sum, variances),}
                                    \DataTypeTok{scale.trees =} \OtherTok{TRUE}\NormalTok{)}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(simulated\_disparity, }\DataTypeTok{main =} \StringTok{"simulated disparity"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(change\_in\_size, }\DataTypeTok{main =} \StringTok{"observed disparity"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Creating a random mass extinction}
\NormalTok{random\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}\DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
                                 \DataTypeTok{condition =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{25}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{),}
                                 \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.75}\NormalTok{))}
\CommentTok{\#\# Creating an extinction that removes species with negative trait values}
\NormalTok{negative\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}\DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
                                   \DataTypeTok{condition =} \KeywordTok{taxa.condition}\NormalTok{(}\DecValTok{25}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{),}
                                   \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DataTypeTok{x =} \DecValTok{0}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Simulate the tree and traits with a random extinction event}
\NormalTok{sim\_rand\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}
                   \DataTypeTok{traits     =}\NormalTok{ my\_traits,}
                   \DataTypeTok{bd.params  =}\NormalTok{ est\_params,}
                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
                   \DataTypeTok{events     =}\NormalTok{ random\_extinction,}
                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
                   \DataTypeTok{replicates =} \DecValTok{50}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Building the tree:....Done.
## Building the tree:.......Done.
## Building the tree:...Done.
## Building the tree:......Done.
## Building the tree:..Done.
## Building the tree:..Done.
## Building the tree:.Done.
## Building the tree:........Done.
## Building the tree:..Done.
## Building the tree:.....Done.
## Building the tree:....Done.
## Building the tree:....Done.
## Building the tree:..Done.
## Building the tree:.................Done.
## Building the tree:....Done.
## Building the tree:....Done.
## Building the tree:..Done.
## Building the tree:......Done.
## Building the tree:.......Done.
## Building the tree:.Done.
## Building the tree:.Done.
## Building the tree:..Done.
## Building the tree:......Done.
## Building the tree:...........Done.
## Building the tree:......Done.
## Building the tree:....Done.
## Building the tree:.Done.
## Building the tree:..................Done.
## Building the tree:....Done.
## Building the tree:...............Done.
## Building the tree:....Done.
## Building the tree:..Done.
## Building the tree:........Done.
## Building the tree:....Done.
## Building the tree:........Done.
## Building the tree:...Done.
## Building the tree:...Done.
## Building the tree:.Done.
## Building the tree:...Done.
## Building the tree:..............Done.
## Building the tree:...Done.
## Building the tree:.Done.
## Building the tree:..Done.
## Building the tree:....Done.
## Building the tree:..Done.
## Building the tree:..Done.
## Building the tree:.Done.
## Building the tree:.Done.
## Building the tree:....Done.
## Building the tree:..........Done.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Simulate the tree and traits with a random extinction event}
\NormalTok{sim\_trait\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}
                   \DataTypeTok{traits     =}\NormalTok{ my\_traits,}
                   \DataTypeTok{bd.params  =}\NormalTok{ est\_params,}
                   \DataTypeTok{stop.rule  =}\NormalTok{ stop\_rule,}
                   \DataTypeTok{events     =}\NormalTok{ negative\_extinction,}
                   \DataTypeTok{null.error =} \DecValTok{100}\NormalTok{,}
                   \DataTypeTok{replicates =} \DecValTok{50}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Building the tree:.....................Done.
## Building the tree:...................Done.
## Building the tree:.....Done.
## Building the tree:.Done.
## Building the tree:....Done.
## Building the tree:..Done.
## Building the tree:..Done.
## Building the tree:......................Done.
## Building the tree:..Done.
## Building the tree:..Done.
## Building the tree:...Done.
## Building the tree:.Done.
## Building the tree:..Done.
## Building the tree:......Done.
## Building the tree:.Done.
## Building the tree:......Done.
## Building the tree:..Done.
## Building the tree:.Done.
## Building the tree:....Done.
## Building the tree:...Done.
## Building the tree:..Done.
## Building the tree:......Done.
## Building the tree:....................Done.
## Building the tree:...Done.
## Building the tree:.........Done.
## Building the tree:.Done.
## Building the tree:...Done.
## Building the tree:.Done.
## Building the tree:.........Done.
## Building the tree:.Done.
## Building the tree:.............Done.
## Building the tree:..............Done.
## Building the tree:......Done.
## Building the tree:.Done.
## Building the tree:.......Done.
## Building the tree:...Done.
## Building the tree:..Done.
## Building the tree:......Done.
## Building the tree:.Done.
## Building the tree:..Done.
## Building the tree:............Done.
## Building the tree:.....Done.
## Building the tree:......Done.
## Building the tree:.Done.
## Building the tree:.Done.
## Building the tree:..............Done.
## Building the tree:........Done.
## Building the tree:..............Done.
## Building the tree:.Done.
## Building the tree:..Done.
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Visualising the difference between both scenarios}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(sim\_rand\_extinction[[}\DecValTok{1}\NormalTok{]], }\DataTypeTok{main =} \StringTok{"Random extinction"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(sim\_trait\_extinction[[}\DecValTok{1}\NormalTok{]], }\DataTypeTok{main =} \StringTok{"Negative extinction"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

Interestingly, with all our parameters used here and our specific disparity metric we cannot detect the effect of a mass extinction.

@@

\hypertarget{quick-overview}{%
\subsection{Quick overview}\label{quick-overview}}

\begin{longtable}[]{@{}llll@{}}
\toprule
\begin{minipage}[b]{0.19\columnwidth}\raggedright
function\strut
\end{minipage} & \begin{minipage}[b]{0.23\columnwidth}\raggedright
arguments\strut
\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\raggedright
input\strut
\end{minipage} & \begin{minipage}[b]{0.33\columnwidth}\raggedright
what does it do\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.19\columnwidth}\raggedright
\texttt{make.traits}\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
process*\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
a function\strut
\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
simulates the trait (see below)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright
n\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
a numeric value\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
the number of dimensions\strut
\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright
start\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
one or more numeric values\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
the starting values for each dimensions\strut
\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright
process.args\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
\ldots{}\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
anything to be passed to \texttt{process} (see below)\strut
\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright
trait.names\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
a character string\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
the name(s) of the trait(s)\strut
\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright
add\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
a \texttt{treats} object\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
another trait generated by \texttt{make.traits}\strut
\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright
update\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
a \texttt{treats} object\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
another trait generated by \texttt{make.traits}\strut
\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright
test\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
logical\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
whether to test the validity of the trait\strut
\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright
background\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
a \texttt{treats} object\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
another trait generated by \texttt{make.traits} (to run in the background)\strut
\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright
---------\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
-----------\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
-------\strut
\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
----------------\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright
\texttt{process}\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
x0\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
a numeric value\strut
\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
the value of the previous trait\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright
edge.length\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
a numeric value\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
the branch length value\strut
\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright
\ldots{}\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
\ldots{}\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
any other argument to be handled by the process\strut
\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

* non-optional arguments

@@

\begin{verbatim}
## $main
## $main$A
## $main$A$process
## function (x0 = 0, edge.length = 1, Sigma = diag(length(x0)), 
##     ...) 
## {
##     return(t(MASS::mvrnorm(n = 1, mu = x0, Sigma = Sigma * edge.length, 
##         ...)))
## }
## <bytecode: 0x5647036c6610>
## <environment: namespace:treats>
## 
## $main$A$start
## [1] 0
## 
## $main$A$trait_id
## [1] 1
## 
## 
## 
## $background
## NULL
\end{verbatim}

@@

\hypertarget{quick-overview-1}{%
\subsection{Quick overview}\label{quick-overview-1}}

\begin{longtable}[]{@{}llll@{}}
\toprule
\begin{minipage}[b]{0.19\columnwidth}\raggedright
function\strut
\end{minipage} & \begin{minipage}[b]{0.23\columnwidth}\raggedright
arguments\strut
\end{minipage} & \begin{minipage}[b]{0.14\columnwidth}\raggedright
input\strut
\end{minipage} & \begin{minipage}[b]{0.33\columnwidth}\raggedright
what does it do\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.19\columnwidth}\raggedright
\texttt{make.modifiers}\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
branch.length\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
a function\strut
\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
generating branch length (see below)\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright
selection\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
a function\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
selecting a lineage (see below)\strut
\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright
speciation\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
a function\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
choosing between speciation and extinction (see below)\strut
\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright
condition\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
a function\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
a function that affects the condition of when to trigger the functions above (see below)\strut
\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright
modify\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
a function\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
a function that modifies the value of the trigger value generated when the \texttt{condition} is met (see below)\strut
\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright
add\strut
\end{minipage} & \begin{minipage}[t]{0.23\columnwidth}\raggedright
a \texttt{treats} object\strut
\end{minipage} & \begin{minipage}[t]{0.14\columnwidth}\raggedright
another modifier generated by \texttt{make.modifiers}\strut
\end{minipage} & \begin{minipage}[t]{0.33\columnwidth}\raggedright
\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.19\columnwidth}\raggedright
